\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{geometry}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{enumitem}
\geometry{a4paper, margin=1in}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolor}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolor},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}

\lstset{style=mystyle}

\title{Basic Probability and Its Applications in AI}
\author{Bui Van Tai}
\date{}
\begin{document}

\maketitle
In this section, we'll examine several examples of probability applied to different types of data.

\section*{3.1 Naive Bayes Classifier for Discrete Random Variables}
\subsection*{Problem Example: Predicting Whether to Play Tennis}

We are given a small dataset to predict whether a person will play tennis based on four \textbf{discrete features} (a "naive" hypothesis, yet effective): \textit{Outlook}, \textit{Temperature}, \textit{Humidity} and \textit{Wind}.

\subsection*{Dataset (5 Samples)}

\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
Outlook & Temp & Humidity & Wind & Play \\
\hline
Sunny & Hot & High & Weak & No \\
Overcast & Hot & High & Weak & Yes \\
Rain & Mild & High & Weak & Yes \\
Sunny & Cool & Normal & Weak & Yes \\
Rain & Mild & High & Strong & No \\
Sunny & Cool & High & Strong & ??? \\
\hline
\end{tabular}
\end{center}

\subsection*{Step 1: Bayes's Theorem Overview}
We want to compute:
$$
P(\text{Play=Yes} \mid X), \quad P(\text{Play=No} \mid X)
$$
where $$X = (\text{Outlook = Sunny}, \text{Temp = Cool}, \text{Humidity = High}, \text{Wind = Strong})$$
According to Bayes's Theorem:

$$
P(\text{Class} \mid X) = \frac{P(X \mid \text{Class}) \cdot P(\text{Class})}{P(X)}
$$
Since $P(X)$ is constant for both classes, we compare:
$$
P(X \mid \text{Yes}) \cdot P(\text{Yes}) \quad \text{and} \quad P(X \mid \text{No}) \cdot P(\text{No})
$$

\subsection*{Step 1: Compute Prior Probabilites}
From the dataset: 

$$
P(\text{Yes}) = \frac{3}{5} = 0.6
\quad , \quad
P(\text{No}) = \frac{2}{5} = 0.4
$$

\subsection*{Step 2: Likelihoods (with Laplace Smoothing)}

Apply Laplace smoothing (add-one) to avoid zero probabilities
$$
P(x_i \mid \text{Class}) = \frac{\text{count}(x_i, \text{Class}) + 1}{\text{count}(\text{Class}) + N_i}
$$

Where $N_i$ is the number of possible values for feature $i$.

- Outlook: 3 values (Sunny, Overcast, Rain)

- Temperature: 3 values (Hot, Mild, Cool)

- Humidity: 2 values (High, Normal)

- Wind: 2 values (Weak, Strong)

\subsection*{Step 3: Compute Likelihoods}

\subsubsection*{For Class = Yes (3 samples)}

$$
P(X \mid \text{Yes}) = P(\text{Sunny} \mid \text{Yes}) \cdot P(\text{Cool} \mid \text{Yes}) \cdot P(\text{High} \mid \text{Yes}) \cdot P(\text{Strong} \mid \text{Yes})
$$

$$
= \frac{1+1}{3+3} \cdot \frac{1+1}{3+3} \cdot \frac{2+1}{3+2} \cdot \frac{0+1}{3+2}
= \frac{2}{6} \cdot \frac{2}{6} \cdot \frac{3}{5} \cdot \frac{1}{5} = \frac{12}{900}
$$

$$
P(X \mid \text{Yes}) \cdot P(\text{Yes}) = \frac{12}{900} \cdot 0.6 = \frac{7.2}{900} \approx 0.008
$$

\subsubsection*{For Class = No (2 samples)}

$$
P(X \mid \text{No}) = P(\text{Sunny} \mid \text{No}) \cdot P(\text{Cool} \mid \text{No}) \cdot P(\text{High} \mid \text{No}) \cdot P(\text{Strong} \mid \text{No})
$$

$$
= \frac{1+1}{2+3} \cdot \frac{0+1}{2+3} \cdot \frac{2+1}{2+2} \cdot \frac{1+1}{2+2}
= \frac{2}{5} \cdot \frac{1}{5} \cdot \frac{3}{4} \cdot \frac{2}{4} = \frac{12}{400}
$$

$$
P(X \mid \text{No}) \cdot P(\text{No}) = \frac{12}{400} \cdot 0.4 = \frac{4.8}{400} = 0.012
$$

\subsection*{Conclusion}

Since:
$$
P(\text{No} \mid X) > P(\text{Yes} \mid X)
$$
We conclude that the person will \textbf{not play tennis}.

\section*{Python Code}
\begin{lstlisting}[language=Python, caption={Naive Bayes Classifier}]
# Training data (5 samples)
data = [
    ['Sunny', 'Hot', 'High', 'Weak', 'No'],
    ['Overcast', 'Hot', 'High', 'Weak', 'Yes'],
    ['Rain', 'Mild', 'High', 'Weak', 'Yes'],
    ['Sunny', 'Cool', 'Normal', 'Weak', 'Yes'],
    ['Rain', 'Mild', 'High', 'Strong', 'No']
]

# Input to predict
X = ['Sunny', 'Cool', 'High', 'Strong']

# Extract labels and features
labels = [row[-1] for row in data]
features = [row[:-1] for row in data]

# Calculate prior probability P(Class)
def prior_prob(class_label):
    return sum(1 for label in labels if label == class_label) / len(labels)

# Calculate conditional probability P(x_i | Class) with Laplace smoothing
def cond_prob(feature_idx, feature_val, class_label):
    count = 0
    total = 0
    unique_vals = set(row[feature_idx] for row in features)

    for i, row in enumerate(features):
        if labels[i] == class_label:
            total += 1
            if row[feature_idx] == feature_val:
                count += 1

    # Apply Laplace smoothing
    return (count + 1) / (total + len(unique_vals))

# Compute P(X | Yes) * P(Yes)
yes_prob = prior_prob('Yes')
for i in range(len(X)):
    yes_prob *= cond_prob(i, X[i], 'Yes')

# Compute P(X | No) * P(No)
no_prob = prior_prob('No')
for i in range(len(X)):
    no_prob *= cond_prob(i, X[i], 'No')

# Final prediction
print(f"P(X | Yes) * P(Yes) = {yes_prob}")
print(f"P(X | No) * P(No) = {no_prob}")
if yes_prob > no_prob:
    print("=> Prediction: Play")
else:
    print("=> Prediction: Do not play")
\end{lstlisting}


$Output$

\begin{lstlisting}
P(X | Yes) * P(Yes) = 0.008
P(X | No) * P(No) = 0.012
=> Prediction: Do not play
\end{lstlisting}


\section*{3.2 Naive Bayes for Continuous Random Variable}

Unlike discrete variables, which can take only a finite or countable set of distinct values, \textit{continuous variables} can take infinite number of possible values (height, weight, temperature,...).

This example shows how to apply Navie Bayes Classifier to continuous data using Gaussian distribution.

We are given 5 BMI values and their corresponding class labels:

\begin{center}
\begin{tabular}{|c|c|}
\hline
BMI (x) & Class (y) \\
\hline
18 & 0 (Healthy) \\
20 & 0 \\
22 & 0 \\
26 & 1 (Sick) \\
28 & 1 \\
24 & ??? \\
\hline
\end{tabular}
\end{center}

We want to classify a new sample with BMI=24.

\subsection*{Step 1: Compute mean and variance for each class}
\subsubsection*{For class $C_0$ (Healthy):}

General formulas:
$$
\mu _0 = \frac{1}{n_0} \sum_{i=1}^{n_0} x_i
\quad  and \quad 
\sigma_0^2 = \frac{1}{n_0} \sum_{i=1}^{n_0} (x_i - \mu  _0)^2
$$

Apply to data: [18, 20, 22]

$$
\mu_0 = \frac{18 + 20 + 22}{3} = 20
\quad
$$
$$
\sigma _0^2 = \frac{(18 - 20)^2 + (20 - 20)^2 + (22 - 20)^2}{3} = \frac{8}{3} \approx 2.67
\quad
$$


$$
\sigma_0 = \sqrt{2.67} \approx 1.63
$$

\subsubsection*{For Class $ C_1 $ (Sick):}
Apply to data: [26, 28]

$$
\mu _1 = \frac{26 + 28}{2} = 27
\quad , \quad
\sigma _1^2 = \frac{(26 - 27)^2 + (28 - 27)^2}{2} = 1
\quad , \quad
\sigma _1 = \sqrt{1} = 1
$$

\subsection*{Step 2: Compute Pior Probabilities}

$$
P(C_0) = \frac{n_0}{n} = \frac{3}{5} = 0.6
$$

$$
P(C_1) = \frac{n_1}{n} = \frac{2}{5} = 0.4
$$

\subsection*{Step 3: Use Gaussian Probability Density Function}

General formula:
$$
P(x|C_k) = \frac{1}{\sqrt{2\pi \sigma_k^2}} \cdot e^{-\frac{(x - \mu_k)^2}{2\sigma_k^2}}
$$

\subsubsection*{For class 0:}

$$
P(x=24|C_0) = \frac{1}{\sqrt{2\pi \cdot 2.67}} \cdot e^{-\frac{(24 - 20)^2}{2 \cdot 2.67}} \approx 0.0122
$$

$$
P(C_0|x=24) \propto 0.0122 \cdot 0.6 = 0.0073
$$


\subsubsection*{For class 1:}
$$
P(x=24|C_1) = \frac{1}{\sqrt{2\pi \cdot 1}} \cdot e^{-\frac{(24 - 27)^2}{2 \cdot 1}} \approx 0.0044
$$

$$
P(C_1|x=24) \propto 0.0044 \cdot 0.4 = 0.0018
$$

\subsection*{Step 4: Final Prediction}

$$
P(C_0|x=24) \propto 0.0073 \quad
P(C_1|x=24) \propto 0.0018
\Rightarrow \text{Predict Class 0 (Healthy)}
$$

\section*{Python Code}
We use scikit-learn library, which provides efficient and easy-to-use tools for implementing Naive Bayes Classifier models, making the process simpler.
\begin{lstlisting}[language=Python, caption={NBC For Continuous Random Variable}]

from sklearn.naive_bayes import GaussianNB
import numpy as np

# Input data
X = np.array([[18], [20], [22], [26], [28]])
y = np.array([0, 0, 0, 1, 1])

# Train model
model = GaussianNB()
model.fit(X, y)

# Predict for BMI = 24
x_test = np.array([[24]])
predicted_class = model.predict(x_test)
print("Predicted class:", predicted_class[0])

\end{lstlisting}

\begin{lstlisting}
Predicted class: 0 
\end{lstlisting}


Bayes'theorem demonstrates its wide-ranging and adaptable applicability in various problem domains and data types.


\section*{Discussion and Conclusion}

The results of the experiments demonstrate that the Naive Bayes classifier, despite its simplicity, 
can make predictions that are intuitively logical. For example, in the tennis scenario, 
the model predicted that the person would not play tennis when the weather conditions were "Sunny, Cool, High humidity, and Strong wind". 
This aligns with common sense: if the weather is too extreme (e.g., strong wind or high humidity),
 it is less likely someone would want to play tennis.

However, while the model gives reasonable results, it also has clear limitations. 
The most important one lies in its core assumption: Naive Bayes treats all features as \textbf{conditionally independent given the class}. 
In reality, weather features such as temperature and humidity, or wind and outlook, are often correlated. 
This assumption prevents the model from capturing interactions between features — 
Ignoring such dependencies may oversimplify the model and reduce accuracy in more complex datasets.

Despite these limitations, Naive Bayes remains a valuable tool, especially for classification tasks where speed and simplicity are prioritized. 
Its interpretability makes it ideal for introductory use in artificial intelligence and for baseline comparisons in real-world applications.

\end{document}